---
title: "Tinkering Around Tidy Tuesday Pt. 2"
author: "Benjamin Egan"
output: html_document
---

#### Created 4/21/2025

```{r load-packages, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse) 
library(ggplot2)
library(skimr)
library(corrr)


```


# Tidy Tuesday - 4/20/2025

The Tidy Tuesday for this week explored fatal crashes on 4/20, the "marijuana holiday." Through the data I'm going to attempt to show that there is actually no link between crashes and inhibition due to marijuana. This should be similar to lab 7's visualization of bad data lab.

```{r setup}

tuesdata <- tidytuesdayR::tt_load(2025, week = 16)

daily_accidents <- tuesdata$daily_accidents
daily_accidents_420 <- tuesdata$daily_accidents_420

view(daily_accidents_420)

```

### The Questions
1. Can you detect any correlations between fatal car crashes and particular days of the year? <br/>
2. What are the most dangerous days of the year for fatal car crashes in the United States? <br/>
3. Is driving on April 20th more dangerous than other days of the year? <br/>
4. How about other holidays, such as Christmas or the 4th of July?


### 1. Correlation between fatal car crashes and particular days of the year

To see if there is a correlation, I'll use the daily_accidents dataset. My first step is to break out the date into separate day, month, and year variables.

```{r separate out date}

daily_accidents <- data.frame(date = daily_accidents$date,
                 year = as.numeric(format(daily_accidents$date, format = "%Y")),
                 month = as.numeric(format(daily_accidents$date, format = "%m")),
                 day = as.numeric(format(daily_accidents$date, format = "%d")),
                 daily_accidents$fatalities_count)

names(daily_accidents)[names(daily_accidents) == "daily_accidents.fatalities_count"] <- "fatalities"

head(daily_accidents)

```

Now that I've broken out the date into it's components, I want to see if fatalities correlate to either the day , the month, year, or specific day.

```{r correlation dates to fatalities}

day_corr <- cor(daily_accidents$day, daily_accidents$fatalities)

month_corr <- cor(daily_accidents$month, daily_accidents$fatalities)

year_corr <- cor(daily_accidents$year, daily_accidents$fatalities)

dates_corr <- tibble(day_corr, month_corr, year_corr)

dates_corr
```

After doing some basic correlations, it looks like month and year have decent correlations around .25, but day has almost no correlation. But what does this mean? I'm going to visualize the distribution to see if we can see a trend.

#### Months

I'll start by examining months.

```{r month corr visual}

daily_accidents_visual <- daily_accidents %>%
  mutate(month_cat = case_when(
    month == 1 ~ "January",
    month == 2 ~ "February",
    month == 3 ~ "March",
    month == 4 ~ "April",
    month == 5 ~ "May",
    month == 6 ~ "June",
    month == 7 ~ "July",
    month == 8 ~ "August",
    month == 9 ~ "September",
    month == 10 ~ "October",
    month == 11 ~ "November",
    month == 12 ~ "December"
  ))

month_regression <- lm(fatalities ~ month, data = daily_accidents_visual)


month_r_values <- tibble(
  month = 1:12,
  month_cat = factor(month.name, levels = month.name),
  pred = predict(month_regression, newdata = tibble(month = 1:12))
)

daily_accidents_visual %>%
  ggplot(aes(
    x = fct(month_cat),
    y = fatalities
  ))+
  geom_jitter(alpha = .4, width = .25)+
  geom_line(data = month_r_values, 
            aes(x = month_cat, y = pred, group = 1), 
            color = "blue", 
            size = 1) +
  theme_bw()+
  labs(
    x = "Month",
    y = "Number of Fatalities in the month"
  )

```

So based on my knowledge of correlation, the .2 means that there is a slight increase in fatalities as we get into the calendar year. It looks more sinusoidal, so I'll let the line follow a more natural curve.

```{r month corr visual 2}


monthly_means <- daily_accidents_visual %>%
  group_by(month_cat) %>%
  summarise(mean_fatalities = mean(fatalities, na.rm = TRUE)) %>%
  mutate(month_cat = factor(month_cat, levels = month.name))

daily_accidents_visual %>%
  ggplot(aes(
    x = fct(month_cat),
    y = fatalities
  ))+
  geom_jitter(alpha = .4, width = .25, color = "gray")+
  geom_smooth(aes(x = month), 
              method = "loess", 
              se = FALSE, 
              color = "blue", 
              size = 1, 
              data = daily_accidents_visual) +
  geom_point(data = monthly_means, 
             aes(x = month_cat, y = mean_fatalities), 
             color = "tomato1", 
             size = 2) +
  theme_bw()+
  labs(
    x = "Month",
    y = "Number of Fatalities in the month",
    title = "Month with non-linear line"
  )

```

As you can see, there is actually more of a sinusoidal curve to the data, as the peak months are late summer/early fall.

#### Days

Even though days had an almost zero correaltion, I'm still interested in seeing how they turn out.

```{r day corr visual}

daily_means <- daily_accidents_visual %>%
  group_by(day) %>%
  summarise(mean_fatalities = mean(fatalities, na.rm = TRUE))

daily_accidents_visual %>%
  ggplot(aes(
    x = day,
    y = fatalities
  ))+
  geom_jitter(alpha = .4, width = .25, color = "gray")+
  geom_point(data = daily_means, aes(x = day, y = mean_fatalities), color = "blue", size = 2) +
  geom_smooth(color = "black", alpha = .6)+
  theme_bw()+
  labs(
    x = "Days",
    y = "Number of Fatalities"
  )

```

There doesn't appear to be any variation at all in the days, which is why the correlation is so low. What is interesting is that there is again a kind of wave pattern that appears roughly every 5-7 days. I would say this corresponded to days of the week, but it would be different for every month and every year.


#### Years

```{r corr year}

daily_accidents_visual %>%
  ggplot(aes(
    x = year,
    y = fatalities
  ))+
  geom_jitter(alpha = .4, width = .25, color = "gray")+
  geom_smooth(color = "black", alpha = .6)+
  theme_bw()+
  labs(
    x = "Year",
    y = "Number of Fatalities"
  )

min(daily_accidents_visual$year)
max(daily_accidents_visual$year)

```

Now this provides a much more interesting look into car fatalities. We have data from 1992 to 2016. There was a massive dip in fatalities around 2005, but an increase around 2013/2014. After some quick googling, it looks like there were behavioral and vehicle safety programs created and implemented around this time that led to the decline. Very cool, and why there was a negative correlation.

### 2. What are the most dangerous days of the year for fatal car crashes in the United States?

```{r most fatal day}

daily_accidents %>%
  arrange(desc(fatalities)) %>%
head()

```

The most fatal day for car crashes was November 3rd, 2007. There were 299 fatal car crashes.

```{r fatal days}

daily_accidents %>%
  distinct(day, fatalities) %>%
    group_by(day) %>%
  summarize(total_fatal = sum(fatalities)) %>%
    arrange(desc(total_fatal)) %>%
head()


daily_accidents %>%
  distinct(day, fatalities) %>%
    group_by(day) %>%
  summarize(avg_fatal = mean(fatalities)) %>%
    arrange(desc(avg_fatal)) %>%
head()

```
The day with the most fatalities is the 3rd of the month, with 19,244 fatal car crashes. The third also hold the highest average amount of fatalities, with around 155 fatalities every 3rd of the month. Maybe we should stop driving on the 3rd of the month...


